---
title: "main_analysis"
author: "Tobias"
date: "2023-06-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#this sets wd to root dir
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())


library(tidyverse) # for data playing
library(rstatix) # for piped stats

# for survival analysis
library(survival)
library(ranger)
library(ggfortify)
library(survminer)
library(coxme)
library(survival)
library(gtsummary)
library(rcompanion) # way to easily get letter groupings for pairwise tests
library(DescTools) # for dunnets test

rm(list=ls())
```

## 2023 osmia bioassay 


```{r data survival, echo=FALSE}


survival <- read_csv("input/2023_bioassay_datasheets_combined.csv", skip_empty_rows = TRUE)%>%
  filter(!end_date == "remove")%>%
  mutate(end_date = as.Date(end_date, format= "%Y/%m/%d"),
         start_date= as.Date(start_date, format= "%Y/%m/%d"), # make date a date
         status = if_else(end_status == "dead", 1, 0))%>% # add 0 and 1 status for survival analysis
  mutate_if(is.character, factor)%>% # change other chr to factors
  mutate( # make a column of treatment names with line breaks for better graph labels
      splits = if_else( 
        str_count(treatment, '\\S+') > 2,
        str_c(word(treatment,1,1), word(treatment,2,str_count(treatment,'\\S+')), sep = '\n'),
        treatment
      )
    ) %>%
  mutate(days_to_event = end_date-start_date) %>% # calculate the days in assay until cocoon or death
  mutate(treatment=relevel(treatment, ref="Osmia pollen (control)"),) %>% # make osmia control first level 
  mutate(day = as.factor(day),
         plate_num = as.factor(plate_num))%>%
  mutate(days_to_death = if_else(end_status == "dead", days_to_event, max(days_to_event))) %>% # i need to make a column where survivors live until the last check point, not just to cocoon
  mutate(days_to_death = as.numeric(days_to_death),
         days_to_event = as.numeric(days_to_event))# and finally fix difftime to numeric


#load pollen staining data
p_stain <- read_csv("input/pollen_staining.csv") %>% #read in pollen staining data
  mutate_if(is.character, factor)%>% # change other chr to factors
  drop_na(date_checked) %>% # drop wells with no frass
  mutate(per_empty = ((`0`+`1-49`)/total)*100)
         # per_full = ((`50-99`+`100`)/total)*100,
         # per_zero = (`0`/total)*100) #adjust counts into percents

# combine with survival and treatment data
pollen_survival <- left_join(survival, p_stain)


# then also load pre digestion pollen staining to correct digestion percent
pre_stain <- read_csv("input/pre_digested_pollen.csv") %>% #read in pollen staining data
  mutate_if(is.character, factor)%>% # change other chr to factors
  mutate(day = as.factor(day))%>%
  mutate(per_empty_pre = ((`0`+`1-49`)/total)*100) %>%
         # per_full_pre = ((`50-99`+`100`)/total)*100,
         # per_zero_pre = (`0`/total)*100)%>% #adjust counts into percents
  select(-c(sample_id,treat_num,`0`,`1-49`,`50-99`,`100`,total))

# now merge pre stain with combined data and correct for pre digestion viability
# formula adapted from Understanding pollen specialization in mason bees: a case study of six species
# by Megan K. McAulay· Saf Z. Killingsworth· Jessica R. K. Forrest

combined <- left_join(pollen_survival, pre_stain)%>%
  mutate(per_empty_cor = (((per_empty - per_empty_pre)/(100-per_empty_pre))*100))%>%
  select(-c(per_empty,  #clean up some columns I wont need
            per_empty_pre, 
            `0`,
            `1-49`,
            `50-99`,
            `100`,
            total, 
            date_stained,
            date_checked, 
            person_counting, 
            notes))


```


``` {r pollen staining data, echo = FALSE}



# then merge the two using sample_id


```

now lets just make a quick summary of 
``` {r summary graphs of death}

prop_death <- combined %>%
  ggplot()+
  geom_bar(aes(x=treatment, fill=end_status), position = "fill")+
  ggtitle("break down of death")+
  scale_x_discrete(guide = guide_axis(angle = 0),labels = function(x) str_wrap(x, width = 7))+
  scale_fill_manual(values=c("lightblue4", "red4"))
prop_death


# how long it took each treatment to die
time_to_death <- combined %>%
  filter(end_status == "dead") %>%
  ggplot(aes(x=splits, y = days_to_event, color = splits))+
  geom_point()+
  geom_boxplot()
time_to_death

```

This is a density graph showing the time until cocoon completion (only showing bees that survived). The development time of control bees is overlayed onto each other treatment.


```{r time of development boxplot}


dev_time<- combined %>%
  filter(end_status=="cocoon")%>%
  ggplot(aes(x=treatment, y=days_to_event, color = treatment))+
  geom_boxplot()+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  theme_bw()

dev_time

ggsave(dev_time, file = "output/graphs/test.svg", width = 7, height = 8)


dunntest_dev<-combined %>%
  filter(end_status=="cocoon")%>%
  dunn_test(days_to_event~treatment)%>%
  unite(comparison, c(group1, group2), sep = "-")
dunntest_dev

dev_time_letters = cldList(p.adj ~ comparison, data=dunntest_dev)
dev_time_letters

```
This shows the differences in development time for the larvae that did successfully spin a cocoon


```{r time of dev density dunnets}

only_control<-combined %>%
  filter(end_status=="cocoon", treatment=="Osmia pollen (control)")%>%
  rename(control=treatment)


combined %>%
  filter(end_status=="cocoon")%>%
  ggplot(aes(x=days_to_event))+
  geom_density(aes(fill=treatment), alpha=.4)+
  geom_density(data=only_control, aes(fill=control), alpha =.2)+
  facet_wrap(~treatment)+
  guides(fill="none")+
  theme_bw()

dunnet_dev<-combined %>%
  filter(end_status=="cocoon")%>%
  DunnettTest(data=., days_to_event~treatment)

dunnet_dev


```
However when running a dunnetts test to just compare each treatment against the control, all developed significantly slower except "osmia with high antibiotics"


``` {r kruskal wallis}

combined %>%
  kruskal_test(end_status ~ treatment)


combined %>%
  dunn_test(end_status ~ treatment, p.adjust.method = "holm")

```



no difference in survival across sex
```{r survival analysis KM}
km_sex <- survfit(Surv(days_to_death, status) ~ sex, data=combined)
ggsurvplot(km_sex, data = combined, pval = TRUE)



km_day <- survfit(Surv(days_to_death, status) ~ day, data=combined)
ggsurvplot(km_day, data = combined, pval = TRUE)



# with all trials
km_treatment <- survfit(Surv(days_to_death, status) ~ treatment, data=combined)
summary(km_treatment, times = c(1,5*(1:12)))

gg<-ggsurvplot(km_treatment, data = combined, pval = TRUE,
           legend = "right",
           conf.int = TRUE,
           conf.int.style = "step",
           conf.int.alpha = .09,
           surv.median.line = "hv")


pairwise_survdiff(Surv(days_to_death, status) ~ treatment, data=combined, p.adjust.method = "BH", rho = 0)


```


Now do the same removing day 1

```{r survival analysis KM - days 2and3}
df23<- combined %>%
  filter(day != 1)

km_sex23 <- survfit(Surv(days_to_death, status) ~ sex, data=df23)
ggsurvplot(km_sex23, data = df23, pval = TRUE)



km_day23 <- survfit(Surv(days_to_death, status) ~ day, data=df23)
ggsurvplot(km_day23, data = df23, pval = TRUE)


km_treatment23 <- survfit(Surv(days_to_death, status) ~ treatment, data=df23)
summary(km_treatment23, times = c(1,5*(1:12)))

gg23<-ggsurvplot(km_treatment23, data = df23, pval = TRUE,
           legend = "right",
           conf.int = TRUE,
           conf.int.style = "step",
           conf.int.alpha = .09,
           surv.median.line = "hv")

pairwise_survdiff(Surv(days_to_death, status) ~ treatment, data=df23, p.adjust.method = "BH", rho = 0)


```

```{r survival cox}


  coxph(Surv(days_to_death, status) ~ treatment, data = combined) %>% 
    tbl_regression(exp = TRUE) 

```


```{r cox split by variable}
aa_fit <-aareg(Surv(days_to_death) ~ treatment + sex + stage + day, data = combined)

autoplot(aa_fit)
```




# pollen staining
```{r pollen staining}

#first explore if pollen pre digestion is different across treatments
combined %>%
  ggplot()+
  geom_boxplot(aes(x=source, y=per_empty_cor, color= treatment))

combined %>% 
  kruskal_test(per_empty_cor~treatment)

combined %>% 
  dunn_test(per_empty_cor~treatment)%>%
  unite(comparison, c(group1, group2), sep = "-")%>%
  cldList(p.adj ~ comparison, data = .)
```
Here higher percent empty corrected (y axis) indicated better digestion ability


so antibiotics don't interfere with digestion ability but sterilization or pollen switching does. Re-addition of (some) microbes to pollen provision does not negate these impacts. 


